Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (\emph{Sida fallax})

### 2023-10-31 ms/ms.qmd
addressed TNB's feedback on revision

\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the ratio of photosynthesis in an amphistomatous leaf to that of the same leaf with the same total stomatal conductance but with gas exchange blocked through the upper (adaxial) surface, which we term "pseudohypostomy". We thus estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of broad leaves, should increase carbon gain by reducing the average diffusion pathlength between stomata and chloroplasts, yet paradoxically this seemingly simple adaptation is uncommon in nature and we don't know why. Understanding variation in stomatal traits like amphistomy is imperative because these tiny pores play an outsized ecological role in the global carbon and water cycles [@hetherington_role_2003; @berry_stomata:_2010]. A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through transpiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997]. Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_ecoevolutionary_2021]. Despite the success of optimality modeling in predicting $g_\mathrm{sw}$, the same modeling approach has so far failed to explain the rarity of amphistomatous leaves [@muir_is_2019]. **This gap between theory and observations strongly implies that we remain ignorant about some key benefits and costs associated with stomata.**

@@

The association between high light and amphistomy suggests that 'sun' leaves have the most to gain in terms of increased photosynthesis from having stomata on both surfaces, as @mott_adaptive_1982 hypothesized. @parkhurst_adaptive_1978 proposed quantifying this benefit as 'amphistomy advantage' ($\mathrm{AA}$), which we adopt here with some modification (see Materials and Methods). This hypothesis has never been tested directly by comparing the photosynthetic rate of an amphistomatous leaf to that of an otherwise identical hypostomatous leaf with the same total stomatal conductance under the same conditions. We propose a straightforward method to do this by experimentally creating a pseudohypostomatous leaf with gas exchange blocked through the upper surface (see Materials and Methods). One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], which will often result in lower $g_\mathrm{ias}$ either by increasing the diffusion path length [@parkhurst_adaptive_1978] or making the airspace less porous. A nonmutually exclusive hypothesis is that if sun leaves have a stomatal density ratio closer to 0.5 (same density on each leaf surface), this will confer a greater advantage than an amphistomatous leaf with most stomata on one surface. In other words, amphistomy doesn't make much difference if one leaf surface has few open stomata on it. We therefore predict that sun leaves will have greater $\mathrm{AA}$ possibly because they have thicker leaves and/or $\mathrm{SR}$ closer to 0.5. We actually report $g_{\mathrm{smax,ratio}}$, which is similar to $\mathrm{SR}$ except that it accounts for differences in both stomatal density and size between surfaces.

@@

The authors thank Kasey Barton for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, TM Perez for advice on leaf sectioning, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.

>>>>

\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the log-ratio of photosynthesis in an amphistomatous leaf to that of the same leaf but with gas exchange blocked through the upper (adaxial) surface, which we term "pseudohypostomy". We used humidity to modulate stomatal conductance and thus compare photosynthetic rates at the same total stomatal conductance. We estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of broad leaves, should increase carbon gain by reducing the average diffusion pathlength between stomata and chloroplasts, yet paradoxically this seemingly simple adaptation is uncommon in nature and we don't know why. Understanding variation in stomatal traits like amphistomy is imperative because these tiny pores play an outsized ecological role in the global carbon and water cycles [@hetherington_role_2003; @berry_stomata:_2010]. A widely applicable, accurate representation of how stomata mediate the relationship between CO$_2$ gained through photosynthesis and water lost through transpiration is essential to predict future climate using Earth Systems Models [@jarvis_interpretation_1976; @biggins_model_1987; @collatz_physiological_1991; @leuning_critical_1995; @sellers_modeling_1997]. Optimality models accurately predict the major cause of water loss, stomatal conductance ($g_\mathrm{sw}$), by assuming plants maximize carbon gain minus a cost of water [@cowan_stomatal_1977; @givnish_economy_1986; @medlyn_reconciling_2011; @lin_optimal_2015; @wang_towards_2017; @franks_comparing_2018; @deans_optimization_2020; @franklin_organizing_2020; @wang_theoretical_2020; @harrison_eco-evolutionary_2021]. Despite the success of optimality modeling in predicting $g_\mathrm{sw}$, the same modeling approach has so far failed to explain the rarity of amphistomatous leaves [@muir_is_2019]. **This gap between theory and observations strongly implies that we remain ignorant about some key benefits and costs associated with stomata.**

@@

The association between high light and amphistomy suggests that 'sun' leaves have the most to gain in terms of increased photosynthesis from having stomata on both surfaces, as @mott_adaptive_1982 hypothesized. @parkhurst_adaptive_1978 proposed quantifying this benefit as 'amphistomy advantage' ($\mathrm{AA}$), which we adopt here with some modification (see Materials and Methods). This hypothesis has never been tested directly by comparing the photosynthetic rate of an amphistomatous leaf to that of an otherwise identical hypostomatous leaf with the same total stomatal conductance under the same conditions. We propose a straightforward method to do this by experimentally creating a pseudohypostomatous leaf with gas exchange blocked through the upper surface (see Materials and Methods). We use humidity to modulate stomatal conductance so that amphi- and pseudohypostomatous leaves can be compared at the same total stomatal conductance. One reason that sun leaves might have greater $\mathrm{AA}$ is that they are usually thicker or denser [@poorter_metaanalysis_2019], which will often result in lower $g_\mathrm{ias}$ either by increasing the diffusion path length [@parkhurst_adaptive_1978] or making the airspace less porous. A nonmutually exclusive hypothesis is that if sun leaves have a stomatal density ratio closer to 0.5 (same density on each leaf surface), this will confer a greater advantage than an amphistomatous leaf with most stomata on one surface. In other words, amphistomy doesn't make much difference if one leaf surface has few open stomata on it. We therefore predict that sun leaves will have greater $\mathrm{AA}$ possibly because they have thicker leaves and/or $\mathrm{SR}$ closer to 0.5. We actually report $g_{\mathrm{smax,ratio}}$, which is similar to $\mathrm{SR}$ except that it accounts for differences in both stomatal density and size between surfaces.

@@

The authors thank Kasey Barton and two anonymous reviewers for feedback on an earlier version of this manuscript, the University of Hawaiʻi honors council for guidance to GT, Tawn Speetjens for access to one montane site, TM Perez for advice on leaf sectioning, startup funds from the University of Hawaiʻi, NSF Award 1929167 to CDM, and NSF Award 2307341 to TNB.

### 2023-10-31 ms/ms.tex
addressed TNB's feedback on revision

% toggle line numbers
\usepackage{lineno}
\linenumbers

% toggle spacing
\usepackage{setspace}
% \singlespacing
\doublespacing

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

\textbf{Methods} We developed a new method to quantify ``amphistomy
advantage'', \(\mathrm{AA}\), as the ratio of photosynthesis in an
amphistomatous leaf to that of the same leaf with the same total
stomatal conductance but with gas exchange blocked through the upper
(adaxial) surface, which we term ``pseudohypostomy''. We thus estimated
\(\mathrm{AA}\) and related physiological and anatomical traits in 12
populations, six coastal (open, sunny) and six montane (closed, shaded),
of the indigenous Hawaiian species ʻilima (\emph{Sida fallax}).

@@

The association between high light and amphistomy suggests that `sun'
leaves have the most to gain in terms of increased photosynthesis from
having stomata on both surfaces, as Mott et al. (1982) hypothesized.
Parkhurst (1978) proposed quantifying this benefit as `amphistomy
advantage' (\(\mathrm{AA}\)), which we adopt here with some modification
(see Materials and Methods). This hypothesis has never been tested
directly by comparing the photosynthetic rate of an amphistomatous leaf
to that of an otherwise identical hypostomatous leaf with the same total
stomatal conductance under the same conditions. We propose a
straightforward method to do this by experimentally creating a
pseudohypostomatous leaf with gas exchange blocked through the upper
surface (see Materials and Methods). One reason that sun leaves might
have greater \(\mathrm{AA}\) is that they are usually thicker or denser
(Poorter et al., 2019), which will often result in lower
\(g_\mathrm{ias}\) either by increasing the diffusion path length
(Parkhurst, 1978) or making the airspace less porous. A nonmutually
exclusive hypothesis is that if sun leaves have a stomatal density ratio
closer to 0.5 (same density on each leaf surface), this will confer a
greater advantage than an amphistomatous leaf with most stomata on one
surface. In other words, amphistomy doesn't make much difference if one
leaf surface has few open stomata on it. We therefore predict that sun
leaves will have greater \(\mathrm{AA}\) possibly because they have
thicker leaves and/or \(\mathrm{SR}\) closer to 0.5. We actually report
\(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
that it accounts for differences in both stomatal density and size
between surfaces.

@@

\leavevmode\vadjust pre{\hypertarget{ref-harrison_ecoevolutionary_2021}{}}%
Harrison, S. P., W. Cramer, O. Franklin, I. C. Prentice, H. Wang, Å.
Brännström, H. Boer, et al. 2021.
\href{https://doi.org/10.1111/nph.17558}{Eco‐evolutionary optimality as
a means to improve vegetation and land‐surface models}. \emph{New
Phytologist} 231: 2125--2141.

\leavevmode\vadjust pre{\hypertarget{ref-harwood_understanding_2021}{}}%
Harwood, R., G. Théroux‐Rancourt, and M. M. Barbour. 2021.
\href{https://doi.org/10.1111/pce.14079}{Understanding airspace in
leaves: {3D} anatomy and directional tortuosity}. \emph{Plant, Cell \&
Environment} 44: 2455--2465.

@@

\leavevmode\vadjust pre{\hypertarget{ref-marquez_assessing_2023}{}}%
Márquez, D. A., H. Stuart‐Williams, L. A. Cernusak, and G. D. Farquhar.
2023. \href{https://doi.org/10.1111/nph.18784}{Assessing the
{\textless{}}span style="font-variant:small-caps;"{\textgreater{}} {CO}
\(_{\textrm{2}}\) {\textless{}}/span{\textgreater{}} concentration at
the surface of photosynthetic mesophyll cells}. \emph{New Phytologist}
238: 1446--1460.

\leavevmode\vadjust pre{\hypertarget{ref-marquez_cuticular_2022}{}}%
Márquez, D. A., H. Stuart‐Williams, G. D. Farquhar, and F. A. Busch.
2022. \href{https://doi.org/10.1111/nph.17588}{Cuticular conductance of
adaxial and abaxial leaf surfaces and its relation to minimum leaf
surface conductance}. \emph{New Phytologist} 233: 156--168.

@@

\leavevmode\vadjust pre{\hypertarget{ref-terashima_irradiance_2006}{}}%
Terashima, I., Y. T. Hanba, Y. Tazoe, P. Vyas, and S. Yano. 2006.
Irradiance and phenotype: Comparative eco-development of sun and shade
leaves in relation to photosynthetic {CO}₂ diffusion. \emph{Journal of
Experimental Botany} 57: 343--354.

>>>>

% toggle line numbers
\usepackage{lineno}
% \linenumbers

% toggle spacing
\usepackage{setspace}
\singlespacing
% \doublespacing

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

@@

\textbf{Methods} We developed a new method to quantify ``amphistomy
advantage'', \(\mathrm{AA}\), as the log-ratio of photosynthesis in an
amphistomatous leaf to that of the same leaf but with gas exchange
blocked through the upper (adaxial) surface, which we term
``pseudohypostomy''. We used humidity to modulate stomatal conductance
and thus compare photosynthetic rates at the same total stomatal
conductance. We estimated \(\mathrm{AA}\) and related physiological and
anatomical traits in 12 populations, six coastal (open, sunny) and six
montane (closed, shaded), of the indigenous Hawaiian species ʻilima
(\emph{Sida fallax}).

@@

The association between high light and amphistomy suggests that `sun'
leaves have the most to gain in terms of increased photosynthesis from
having stomata on both surfaces, as Mott et al. (1982) hypothesized.
Parkhurst (1978) proposed quantifying this benefit as `amphistomy
advantage' (\(\mathrm{AA}\)), which we adopt here with some modification
(see Materials and Methods). This hypothesis has never been tested
directly by comparing the photosynthetic rate of an amphistomatous leaf
to that of an otherwise identical hypostomatous leaf with the same total
stomatal conductance under the same conditions. We propose a
straightforward method to do this by experimentally creating a
pseudohypostomatous leaf with gas exchange blocked through the upper
surface (see Materials and Methods). We use humidity to modulate
stomatal conductance so that amphi- and pseudohypostomatous leaves can
be compared at the same total stomatal conductance. One reason that sun
leaves might have greater \(\mathrm{AA}\) is that they are usually
thicker or denser (Poorter et al., 2019), which will often result in
lower \(g_\mathrm{ias}\) either by increasing the diffusion path length
(Parkhurst, 1978) or making the airspace less porous. A nonmutually
exclusive hypothesis is that if sun leaves have a stomatal density ratio
closer to 0.5 (same density on each leaf surface), this will confer a
greater advantage than an amphistomatous leaf with most stomata on one
surface. In other words, amphistomy doesn't make much difference if one
leaf surface has few open stomata on it. We therefore predict that sun
leaves will have greater \(\mathrm{AA}\) possibly because they have
thicker leaves and/or \(\mathrm{SR}\) closer to 0.5. We actually report
\(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
that it accounts for differences in both stomatal density and size
between surfaces.

@@

\leavevmode\vadjust pre{\hypertarget{ref-harrison_eco-evolutionary_2021}{}}%
Harrison, S. P., W. Cramer, O. Franklin, I. C. Prentice, H. Wang, Å.
Brännström, H. Boer, et al. 2021.
\href{https://doi.org/10.1111/nph.17558}{Eco-evolutionary optimality as
a means to improve vegetation and land-surface models}. \emph{New
Phytologist} 231: 2125--2141.

\leavevmode\vadjust pre{\hypertarget{ref-harwood_understanding_2021}{}}%
Harwood, R., G. Théroux-Rancourt, and M. M. Barbour. 2021.
\href{https://doi.org/10.1111/pce.14079}{Understanding airspace in
leaves: {3D} anatomy and directional tortuosity}. \emph{Plant, Cell \&
Environment} 44: 2455--2465.

@@

\leavevmode\vadjust pre{\hypertarget{ref-marquez_assessing_2023}{}}%
Márquez, D. A., H. Stuart-Williams, L. A. Cernusak, and G. D. Farquhar.
2023. \href{https://doi.org/10.1111/nph.18784}{Assessing the
{CO}\(_{\textrm{2}}\) concentration at the surface of photosynthetic
mesophyll cells}. \emph{New Phytologist} 238: 1446--1460.

\leavevmode\vadjust pre{\hypertarget{ref-marquez_cuticular_2022}{}}%
Márquez, D. A., H. Stuart-Williams, G. D. Farquhar, and F. A. Busch.
2022. \href{https://doi.org/10.1111/nph.17588}{Cuticular conductance of
adaxial and abaxial leaf surfaces and its relation to minimum leaf
surface conductance}. \emph{New Phytologist} 233: 156--168.

@@

\leavevmode\vadjust pre{\hypertarget{ref-terashima_irradiance_2006}{}}%
Terashima, I., Y. T. Hanba, Y. Tazoe, P. Vyas, and S. Yano. 2006.
Irradiance and phenotype: Comparative eco-development of sun and shade
leaves in relation to photosynthetic {CO}\(_{\textrm{2}}\) diffusion.
\emph{Journal of Experimental Botany} 57: 343--354.
